Manuscript results: Forecasting influenza using ARIMA and ARIMAX

### 2025-05-21 manuscript_results.Rmd
update

# Plot
plt0 <- ggplot(df_combined_, aes(x = Julian_date, y = WIS, group = STATE)) + 
  geom_vline(data = highlight_dates, aes(xintercept = Julian_date), color = "red", linetype = "dashed", size = 1) +  
  geom_line(color = "black", size = 0.1) +  
  facet_wrap(~ Dataset, ncol = 2, scales = "free_y") +  
  theme_minimal() +  
  labs(title = "Figure 2: Baseline model WIS for 48 states on the contiguous U.S.",#,
       subtitle = "Weeks between the dashed red lines are outside of the influenza season",
       x = "", 
       y = "WIS") +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 16, face = "bold"),
        legend.position = "none") +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b %y"
  ) # Format x-axis to show months and set monthly breaks

@@

  # Create the box plot with additional facet for categories
plt1<-  ggplot(combined_data, aes(x = week, y = value, fill = Models)) +
    geom_boxplot(color = "black") +
    scale_fill_manual(values = custom_colors) +
    facet_wrap(category ~ Models , nrow = 3) +
    labs(
      title = "A) Models without log-back transformations",
      x = "Target Week",
      y = "Mean Weighted Interval Score (WIS)"
    ) +
  scale_y_continuous(limits = c(0, 650)) +  # Set y-axis range
    theme_minimal()
plt1
#ggsave(plt1, height = 8, width = 10, filename="fig1A.jpg")  
```

@@

  # Create the box plot with additional facet for categories
plt2<-  ggplot(combined_data, aes(x = week, y = value, fill = Models)) +
    geom_boxplot(color = "black") +
    scale_fill_manual(values = custom_colors) +
    facet_wrap(category ~ Models , nrow = 3) +
    labs(
      title = "B) Models with log-back transformations",
      x = "",
      y = "Weighted Interval Score (WIS)"
    ) +
  scale_y_continuous(limits = c(0, 650)) +  # Set y-axis range
    theme_minimal()
plt2
#ggsave(plt2, height = 8, width = 10, filename="Fig1B.jpg")  
```

@@

# Combine in a 2x2 grid
combined_plot <- (plt1 | plt2) + 
  plot_annotation(
    title = "Figure 3: Mean WIS for our 24 models",
    subtitle = "Results by target week",
    theme = theme(plot.title = element_text(size = 16, face = "bold"),
                  plot.subtitle = element_text(size = 14))
  )
# Print
print(combined_plot)

# Save to file
ggsave("Fig3.jpg", combined_plot, width =13, height =7)
```

@@

# Create the plot
combined_plot <- ggplot(all_weeks, aes(x = Best_Result, fill = Best_Result)) +
  geom_bar() +
  scale_fill_manual(values = custom_colors) +
  labs(title = "Figure 4: Best model based on the lowest mean WIS",
       subtitle = "Results for the 48 states on the contiguous U.S. by target week",
       x = "", y = "Number of states",
       fill = "Best model" ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 16, face = "bold")) +
  facet_grid(Week_Ahead ~ Model_Type, scales = "free_x")

@@

# Heatmap plot with models ordered by WeekAhead == 1  
plot_m_heatmap <- ggplot(p_df, aes(x = factor(WeekAhead), y = Model, fill = m)) +
  geom_tile(aes(fill = ifelse(Significance == "Significant", m, NA)), color = "black") +  # Fill only significant ones
  geom_tile(data = p_df %>% filter(Significance == "Not Significant"), 
            aes(x = factor(WeekAhead), y = Model), fill = "gray80", color = "black") +  # Gray for non-significant
  geom_text(aes(label = paste0("m:", round(m, 1),", sd:", round(sd, 1))), 
            color = "black", size = 4) +  # Add text labels
  scale_fill_gradient(low = "cyan", high = "red", na.value = "gray80", 
                      name = "Percentage (%)", limits = c(-35, 35)) +  # Fixed color scale from -30 to 30
  labs(title = "Figure 5: Mean differences between models' WIS and the baseline across 48 states",
       subtitle = "Gray boxes indicate models that are not significantly different from the baseline (p>0.05)",
         caption = "m = mean difference (%), sd = standard deviation",
       x = "Target Ahead",
       y = "") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
  plot.caption = element_text(size = 16),
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),) +
  facet_wrap(~ Model_Type, scales = "free_y")  # Separate NoLg and other models

@@

```{r}
# Combine in a 2x2 grid
combined_plot <- (ES_1WEEK | ES_2WEEK) /
                 (ES_3WEEK | ES_4WEEK) + 
  plot_annotation(
    title = "Figure 6: Mean WIS for the 48 states in the contiguous U.S.",
    subtitle = "Results for the the 2022/2023 and 2023/2024 influenza seasons",
    theme = theme(plot.title = element_text(size = 16, face = "bold"),
                  plot.subtitle = element_text(size = 14))
  )
# Print
print(combined_plot)

@@

# Plot
reg2<-ggplot(regression_models, aes(x = Week_Ahead, y = independent_variable, fill = significance)) +  
  geom_tile(color = "black") +  # Add black borders to squares
  geom_text(aes(label = round(r_squared, 3)), color = "black", size = 6) +  # Text inside boxes
  scale_fill_manual(values = c("Significant" = "cyan", "Not Significant" = "gray")) +  
  labs(title = "Figure 7: Regression models",
       subtitle = "Mean WIS as dependent variable",
       x = "Week Ahead",
       y = "Independent Variables",
       fill = "Significance (p < 0.05)") +  
  theme_minimal() +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.text.y = element_text(size = 10),
        plot.title = element_text(size = 16, face = "bold"),
                  plot.subtitle = element_text(size = 14))

@@

# Combine in a 2x2 grid
combined_plot <- (ES_1WEEK | ES_2WEEK) /
                 (ES_3WEEK | ES_4WEEK) + 
  plot_annotation(
    title = "Figure 8: Mean WIS differences (%) compared to the baseline",
    subtitle = "Results by forecast horizon",
    theme = theme(plot.title = element_text(size = 16, face = "bold"),
                  plot.subtitle = element_text(size = 14))
  )
# Print
print(combined_plot)

@@

# Plot
reg2<-ggplot(regression_models, aes(x = Week_Ahead, y = independent_variable, fill = significance)) +  
  geom_tile(color = "black") +  # Add black borders to squares
  geom_text(aes(label = round(r_squared, 3)), color = "black", size = 6) +  # Text inside boxes
  scale_fill_manual(values = c("Significant" = "cyan", "Not Significant" = "gray")) +  
  labs(title = "Figure 9: Regression models",
       subtitle = "Mean WIS differences (%) as dependent variable",
       x = "Target Ahead",
       y = "Independent Variables",
       fill = "Significance (p < 0.05)") +    
  theme_minimal() +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.text.y = element_text(size = 10),
        plot.title = element_text(size = 16, face = "bold"),
                  plot.subtitle = element_text(size = 14))

@@

Now let's evaluate which was the mean WIS improvement by epidemiological week

@@

  # Sum values by Julian date across all states
  summed_data <- data__ %>%
    group_by(Julian_date) %>%
    summarize(
      AUTO_AR = sum(AUTO_AR, na.rm = TRUE),  # Sum across all states
      AUTO_AVG_LB = sum(AUTO_AVG_LB, na.rm = TRUE)           # Sum across all states
    ) %>%
    # Calculate the difference ratio
    mutate(Difference =  ((AUTO_AVG_LB / AUTO_AR)-1)*100,
           Week = paste0("Week ", week))

@@

# Plot the difference using ggplot2
model_improv<-ggplot(summed_data_all, aes(x = Julian_date, y = (Difference), color = Week)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed", size = 0.8) +
  theme_minimal() +
  labs(
    title = "Figure 10: Mean percentage of WIS differences across all states",
    subtitle = "AUTO_AVG_LB compared to baseline",
    x = "",
    y = "Difference (%)"
  ) +
  scale_y_continuous(
    limits = c(-100, 100),
    breaks = seq(-100, 100, by = 20)
  ) +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b %y"
  ) +
  scale_color_manual(
    values = c("Week 1" = "#ffcccc",
               "Week 2" = "#ff6666",
               "Week 3" = "#cc0000",
               "Week 4" = "#990000")
  ) +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

@@

ggsave(model_improv, height = 4.5, width = 9,  filename="Fig10.jpg")  
```
Talvez incluir standar deviation das diferenceas

# Create a list of data frames for the four weeks
weeks <- 1:4
summed_data_list <- list()

  # Calculate the percentage difference at the row level (i.e., per state-date)
  data__ <- data__ %>%
    mutate(Difference = ((AUTO_AVG_LB / AUTO_AR) - 1) * 100)

  # Group by Julian date and calculate mean and standard deviation of Difference
  summarized <- data__ %>%
    group_by(Julian_date) %>%
    summarize(
      Mean_Diff = Difference,
      SD_Diff = sd(Difference, na.rm = TRUE)
    ) %>%
    mutate(Week = paste0("Week ", week))

  # Store the summarized data
  summed_data_list[[week]] <- summarized
}

# Plot the mean difference with error bars for standard deviation
model_improv <- ggplot(summed_data_all, aes(x = Julian_date, y = Mean_Diff, color = Week)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = Mean_Diff - SD_Diff, ymax = Mean_Diff + SD_Diff), width = 0.5) +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed", size = 0.8) +
  theme_minimal() +
  labs(
    title = "Figure 10: Mean percentage of WIS differences across all states",
    subtitle = "AUTO_AVG_LB compared to baseline",
    x = "",
    y = "Difference (%)"
  ) +
  scale_y_continuous(
    limits = c(-100, 100),
    breaks = seq(-100, 100, by = 20)
  ) +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b %y"
  ) +
  scale_color_manual(
    values = c("Week 1" = "#ffcccc",
               "Week 2" = "#ff6666",
               "Week 3" = "#cc0000",
               "Week 4" = "#990000")
  ) +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

#ggsave(model_improv, height = 4.5, width = 9, filename = "Fig11.jpg")

>>>>

# Plot
plt0 <- ggplot(df_combined_, aes(x = Julian_date, y = WIS, group = STATE)) + 
  geom_vline(data = highlight_dates, aes(xintercept = Julian_date), color = "red", linetype = "dashed", size = 1) +  
  geom_line(color = "black", size = 0.1) +  
  facet_wrap(~ Dataset, ncol = 2, scales = "free_y") +  
  theme_minimal() +  
  labs(title = "Baseline model WIS for 48 states on the contiguous U.S.",#,
       subtitle = "Weeks between the dashed red lines are outside of the influenza season",
       x = "", 
       y = "WIS") +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 16, face = "bold"),
        legend.position = "none") +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b %y"
  ) # Format x-axis to show months and set monthly breaks

@@

  # Create the box plot with additional facet for categories
plt1<-  ggplot(combined_data, aes(x = week, y = value, fill = Models)) +
    geom_boxplot(color = "black") +
    scale_fill_manual(values = custom_colors) +
    facet_wrap(category ~ Models , nrow = 5) +
    labs(
      title = "A) Models without log-back transformations",
      x = "",
      y = "Mean Weighted Interval Score (WIS)"
    ) +
  scale_y_continuous(limits = c(0, 650)) +  # Set y-axis range
    theme_minimal()+
  theme(legend.position = "none")
plt1
#ggsave(plt1, height = 12, width = 8, filename="fig1A.jpg")  
```

@@

  # Create the box plot with additional facet for categories
plt2<-  ggplot(combined_data, aes(x = week, y = value, fill = Models)) +
    geom_boxplot(color = "black") +
    scale_fill_manual(values = custom_colors) +
    facet_wrap(category ~ Models , nrow = 5) +
    labs(
      title = "B) Models with log-back transformations",
      x = "",
      y = "Weighted Interval Score (WIS)"
    ) +
  scale_y_continuous(limits = c(0, 650)) +  # Set y-axis range
    theme_minimal()+
  theme(legend.position = "none")
plt2
#ggsave(plt2, height = 8, width = 10, filename="Fig1B.jpg")  
```

@@

# Combine in a 2x2 grid
combined_plot <- (plt1 | plt2) + 
  plot_annotation(
    title = "Mean WIS for the 24 models",
    subtitle = "Results by target week",
    theme = theme(plot.title = element_text(size = 16, face = "bold"),
                  plot.subtitle = element_text(size = 14))
  )
# Print
print(combined_plot)

# Save to file
ggsave("Fig3.jpg", combined_plot, width =9, height =10)
```

@@

# Create the plot
combined_plot <- ggplot(all_weeks, aes(x = Best_Result, fill = Best_Result)) +
  geom_bar() +
  scale_fill_manual(values = custom_colors) +
  labs(title = "Best model based on the lowest mean WIS",
       subtitle = "Results for the 48 states on the contiguous U.S. by target week",
       x = "", y = "Number of states",
       fill = "Best model" ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 16, face = "bold")) +
  facet_grid(Week_Ahead ~ Model_Type, scales = "free_x")

@@

# Heatmap plot with models ordered by WeekAhead == 1  
plot_m_heatmap <- ggplot(p_df, aes(x = factor(WeekAhead), y = Model, fill = m)) +
  geom_tile(aes(fill = ifelse(Significance == "Significant", m, NA)), color = "black") +  # Fill only significant ones
  geom_tile(data = p_df %>% filter(Significance == "Not Significant"), 
            aes(x = factor(WeekAhead), y = Model), fill = "gray80", color = "black") +  # Gray for non-significant
  geom_text(aes(label = paste0("m:", round(m, 1),", sd:", round(sd, 1))), 
            color = "black", size = 4) +  # Add text labels
  scale_fill_gradient(low = "cyan", high = "red", na.value = "gray80", 
                      name = "Percentage (%)", limits = c(-35, 35)) +  # Fixed color scale from -30 to 30
  labs(title = "Mean differences between models' WIS and the baseline across 48 states",
       subtitle = "Gray boxes indicate models that are not significantly different from the baseline (p>0.05)",
         caption = "m = mean difference (%), sd = standard deviation",
       x = "Target Ahead",
       y = "") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
  plot.caption = element_text(size = 16),
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),) +
  facet_wrap(~ Model_Type, scales = "free_y")  # Separate NoLg and other models

@@

```{r}
# Combine in a 2x2 grid
combined_plot <- (ES_1WEEK | ES_2WEEK) /
                 (ES_3WEEK | ES_4WEEK) + 
  plot_annotation(
    title = "Mean WIS for the 48 states in the contiguous U.S.",
    subtitle = "Results for the the 2022/2023 and 2023/2024 influenza seasons",
    theme = theme(plot.title = element_text(size = 16, face = "bold"),
                  plot.subtitle = element_text(size = 14))
  )
# Print
print(combined_plot)

@@

# Plot
reg2<-ggplot(regression_models, aes(x = Week_Ahead, y = independent_variable, fill = significance)) +  
  geom_tile(color = "black") +  # Add black borders to squares
  geom_text(aes(label = round(r_squared, 3)), color = "black", size = 6) +  # Text inside boxes
  scale_fill_manual(values = c("Significant" = "cyan", "Not Significant" = "gray")) +  
  labs(title = "Regression models R²",
       subtitle = "Mean WIS as dependent variable",
       x = "Target Week",
       y = "Independent Variables",
       fill = "Significance (p < 0.05)") +  
  theme_minimal() +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.text.y = element_text(size = 10),
        plot.title = element_text(size = 16, face = "bold"),
                  plot.subtitle = element_text(size = 14))

@@

# Combine in a 2x2 grid
combined_plot <- (ES_1WEEK | ES_2WEEK) /
                 (ES_3WEEK | ES_4WEEK) + 
  plot_annotation(
    title = "Mean WIS differences compared to the baseline",
    subtitle = "Results by forecast horizon",
    theme = theme(plot.title = element_text(size = 16, face = "bold"),
                  plot.subtitle = element_text(size = 14))
  )
# Print
print(combined_plot)

@@

# Plot
reg2<-ggplot(regression_models, aes(x = Week_Ahead, y = independent_variable, fill = significance)) +  
  geom_tile(color = "black") +  # Add black borders to squares
  geom_text(aes(label = round(r_squared, 3)), color = "black", size = 6) +  # Text inside boxes
  scale_fill_manual(values = c("Significant" = "cyan", "Not Significant" = "gray")) +  
  labs(title = "Regression models R²",
       subtitle = "Mean WIS differences (%) as dependent variable",
       x = "Target Week",
       y = "Independent Variables",
       fill = "Significance (p < 0.05)") +    
  theme_minimal() +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.text.y = element_text(size = 10),
        plot.title = element_text(size = 16, face = "bold"),
                  plot.subtitle = element_text(size = 14))

@@

@@

  # Sum values by Julian date across all states
  summed_data <- data__ %>%
    group_by(Julian_date) %>%
    summarize(
      AUTO_AR = mean(AUTO_AR, na.rm = TRUE),  # Sum across all states
      AUTO_AVG_LB = mean(AUTO_AVG_LB, na.rm = TRUE)           # Sum across all states
    ) %>%
    # Calculate the difference ratio
    mutate(Difference =  ((AUTO_AVG_LB / AUTO_AR)-1)*100,
           Week = paste0("Week ", week))

@@

# Plot the difference using ggplot2
model_improv<-ggplot(summed_data_all, aes(x = Julian_date, y = (Difference), color = Week)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed", size = 0.8) +
  theme_minimal() +
  labs(
    title = "Mean percentage of WIS differences across all states",
    subtitle = "AUTO_AVG_LB compared to baseline",
    x = "",
    y = "Difference (%)"
  ) +
  scale_y_continuous(
    limits = c(-100, 100),
    breaks = seq(-100, 100, by = 20)
  ) +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b %y"
  ) +
  scale_color_manual(
    values = c("Week 1" = "#ffcccc",
               "Week 2" = "#ff6666",
               "Week 3" = "#cc0000",
               "Week 4" = "#990000")
  ) +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

@@

ggsave(model_improv, height = 4.5, width = 9,  filename="Fig10.jpg")  
```
